Translation of B implementations to the LLVM (\textsc{draft})

### 2012-11-13 b0-llvm.tex
- Created the example with the new parameter in functions ( it keeps the state of module).
- Added some elements in the abstract tree.

- Adjusted the dictionary to support rather than (Node -> text), support (text -> text) and domain B keeps identifiers and image identifiers LLVM.

- Fix section 6.1, apparently without using two dictionaries. Fix to use only one.

- Added a parameter in functions to get the reference of the global variables.

\begin{table}
  \begin{center}
    \begin{tabular}[t]{cc}
      \begin{tabular}[t]{|l|l|l|}
        \hline
        \textsf{k} & Decription & attributes\\
        \hline
        \hline
        \textsf{Impl} & Implementation & \textsf{sees} : ? \\
        & & \textsf{imports} : ? \\
        & & $\cdots$ \\
        & & \textsf{ops} : Op \\
        \hline
        \textsf{Op} & Operation & \textsf{header}: \textsf{Hd} \\
        & & $\textsf{body} \in \textsf{inst}$ \\
        & & \textsf{next} : \textsf{Op} \\
        \hline
        \textsf{Hd} & Operation header & \textsf{ident} : text \\
        & & \textsf{output} : \textsf{VarDecl} \\
        & & \textsf{input} : \textsf{VarDecl} \\
        \hline
        \textsf{Var} & Variable, parameter & \textsf{ident} : text \\
        & & \textsf{type} : basic type \\
        & & \textsf{next} : \textsf{Var} \\
        \hline
      \end{tabular}
      &
      \begin{tabular}[t]{|l|l|l|}
        \hline
        \textsf{k} & Decription & attributes\\
        \hline
        \hline
        \textsf{Blk} & Block instr. & $\textsf{body} \in \textsf{inst}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{VarDecl} & Variable decl. & \textsf{ident}: \textsf{VarDecl} \\
        & & $\textsf{body} \in \textsf{inst}$ \\
        \hline
        \textsf{If} & If & \textsf{branches}: \textsf{IfBr} \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{IfBr} & If branch & \textsf{cond}: predicate \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & \textsf{next} : \textsf{IfBr} \\
        \hline
        \textsf{BEq} & Becomes equal & \textsf{lhs}: target \\
        & & \textsf{rhs}: expression \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Call} & Call instruction & \textsf{fun}: name \\
        & & \textsf{args}: expression \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{While} & While instr. & \textsf{cond}: predicate \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Assert} & Assert instr. & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Case} & Case instr. & \textsf{expr}: expression\\
        & & \textsf{branches}: \textsf{CaseBr} \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{CaseBr} & Case branch & \textsf{val}: \textsf{Value} \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & \textsf{next} : \textsf{CaseBr} \\
        \hline
        \textsf{Value} & Value & ??? \\
        & & \textsf{next}: \textsf{Value} \\
        \hline
      \end{tabular} \\
      \\
      \multicolumn{2}{c}{Note: $\textsf{Instr} = \{\textsf{Blk}, \textsf{VarDecl}, \textsf{If}, \textsf{BEq}, \textsf{Call}, \textsf{While}, \textsf{Assert}, \textsf{Case}\}$} \\
    \end{tabular}
    \caption{Structure of the parse tree nodes. For each attribute we give the name of the attribute and
    the type or kind(s) of node that it may be.}
    \label{tab:node-attr}
  \end{center}
\end{table}

Code generation is defined through a set of numbered rules. The
translation rules for some element \textsf{E} of the B0 language are
presented using a pseudo-functional language.  In such rules, the
\textsf{sans serif} and \texttt{typewriter} fonts are used to denote
\textsf{B0} abstract syntax tree nodes and \texttt{LLVM} text
respectively, and $\conc$ denotes string catenation and $\nl$ a new
line. .

@@

$\fresh: text \times dict \rightarrow text$ is such that
$\fresh(t, D)$  returns a text $t'$ that is either $t$, if $t$ is not in
the domain of the dictionary $D$, or $t$ followed by the string $n$ of
digits representing the smallest positive number $t \conc n$ does not
belong to the domain of $D$.

@@

We first define the translation rule for the root of the source
program abstract syntax tree, which shall correspond to a B
implementation. This is a preliminary version where the code
generation is only able to handle standalone B0 implementations, i.e.
the state is defined in the ``concrete variables'' clause. Moreover
the data is restricted to predefined basic types \textsf{INT} and
\textsf{BOOL}:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{i.values}{val} \textbf{ and } O_v = \trad{i.vars}{var} \\
  & \quad \textbf{ and } O_{in} = \trad{i.initialization}{ini} \textbf{ and } O_{op} = \trad{i.ops}{ops} \\
  & \quad \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad \trad{i.ops}{ops}  \\
  & \textbf{ where } \textsf{i.k} = \textsf{Impl} 
\end{split}
\end{align}
\end{framed}
Code generation for initialisation and operations is specified in
section~\ref{sec:control}. Code generation for data (type and
state) is defined in section~\ref{sec:data}, and for expression
evaluation in section~\ref{sec:expr}.

@@

% TODO: Qual o valor de n ?  Seria isto: n = L()  (\name) ?
% Nas duas funções seguintes parece ter uma mistura com
% relação ao nome das funções:  {var}, {type}, {val} 
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{var} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{type} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident}  \textbf{ and }  n = ???? \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = internal global "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{type} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

Constant valuations are translated to constants in the
LLVM program, as described in the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{val} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{val} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ and }  n = ???? \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = constant "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{val} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

This section addresses how B operations are translated to LLVM
functions. Each operation is translated in turn, and the result of
the translation of individual operations is concatenated to yield the
result of the translation of the operations clause:
\begin{framed}
\begin{align}
  \begin{split}
  \trad{}{ops} : & \textsf{ Node} \rightarrow text \\
  \trad{o}{ops} \isdef 
  & \textbf{ yields } \trad{o}{op} \conc \trad{o.next}{ops}  \\
  & \textbf{ where } \textsf{o} \neq \bot  \textbf{ and } \textsf{o.k} \in \textsf{Op}  \\
  \trad{$\bot$}{ops} \isdef
  & \textbf{ yields } \mty 
  \end{split}
\end{align}
\end{framed}

@@

The translation of an operation is defined by the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{op} &: \textsf{ Node} \rightarrow text \\
  \trad{op}{op} \isdef
  & \textbf{ let } O_l = \trad{op}{link} \textbf{ and } O_h, D_h = \trad{op.header}{header} \textbf{ in}  \\
  & \textbf{ let } O_a, D_a = \trad{op.body, $D_h$}{alloc} \textbf{ in}  \\
  & \textbf{ let } O_i = \trad{op.body, $D_a$}{inst} \textbf{ in}  \\
  & \textbf{ yields } \texttt{"declare "} \conc O_l \conc O_h \conc \texttt{"\{"} \conc \nl \conc \\
  & \quad \conc O_a \conc \nl \conc O_i \conc \nl \conc \texttt{"ret void"} \conc \nl \conc \texttt{"\}"} \conc \nl  \\
  & \textbf{ where } \textsf{op} \ne \bot \textbf{ and } \textsf{op.k} = \textsf{Op} 
\end{split}
\end{align}
\end{framed}

@@

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{header} &: Hd \rightarrow text \times dict \times dict \\
    \trad{h}{header} \isdef& \textbf{ let } T_n = \textsf{"@"} \conc \textsf{h.ident} 
    \textbf{ and } T_i, D_i = \trad{h.input}{inp} 
    \textbf{ and } T_o, D_o = \trad{h.output}{out} \textbf{ in}  \\
    & \textbf{ let } T_p = T_i \conc^, T_o \textbf{ in}  \\
    & \textbf{ yields } \texttt{"void "} \conc T_n \conc \texttt{"("} \conc T_p \conc
    \texttt{")"}, D_i \cup D_o, \{ \textsf{h.ident} \mapsto T_n \}  \\
    & \textbf{ where } \textsf{h} \ne \bot \textbf{ and } \textsf{h.k} = \textsf{Hd} 
  \end{split}
\end{align}
\end{framed}

@@

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{inp} &: \textsf{ Node } \rightarrow text \times dict \\
    \trad{i}{inp} \isdef& \textbf{ let } T_n = \textsf{"\%"} \conc \textsf{i.ident} 
    \textbf{ and } T_t = \trad{i.type}{type} 
    \textbf{ and } T, D = \trad{i.next}{inp} \textbf{ in}  \\
    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{i} \mapsto T_n \} \cup D  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Var}  \\
    \trad{$\bot$}{inp} \isdef& \textbf{ yields } \mty, \emptyset 
  \end{split}
\end{align}
\end{framed}

@@

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{out} &: \textsf{ Node } \rightarrow text \times dict \\
    \trad{o}{out} \isdef& \textbf{ let } T_n = \textsf{"\%"} \conc \textsf{o.ident} 
    \textbf{ and } T_t = \trad{o.type}{type} \conc \texttt{"*"}
    \textbf{ and } T, D = \trad{o.next}{out} \textbf{ in}  \\
    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{o} \mapsto T_n \} \cup D  \\
    & \textbf{ where } \textsf{o} \ne \bot \textbf{ and } \textsf{o.k} = \textsf{Var}  \\
    \trad{$\bot$}{out} \isdef& \textbf{ yields } \mty, \emptyset  
  \end{split}
\end{align}
\end{framed}

@@

One must avoid generating name conflicts in the LLVM output. For each
B0 variable \textsf{v} found, a fresh LLVM name \texttt{l} is thus
created. This association is recorded in the $D$ parameters and results.
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{alloc}_{base} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ yields } \trad{i.body, $D$}{alloc}  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{Blk}, \textsf{While}, \textsf{IfBr}, \textsf{CaseBr} \}  \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ yields } \trad{i.branch, $D$}{alloc}  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{If}, \textsf{Case} \}  \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ let } O_l, D_l = \trad{i.ident, $D$}{local} \textbf{ in}  \\
    & \textbf{ let } O_b, D_b = \trad{i.body, $D_l$}{alloc} \textbf{ in}  \\
    & \textbf{ yields } O_l \conc \nl \conc O_b, D_b  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{VarDecl} 
  \end{split}
\end{align}
\end{framed}

To illustrate how heap allocation works in LLVM, this is the statement
produced by Clang corresponding to the C declaration
\verb^unsigned short c1^:
\begin{verbatim}
  %c1 = alloca i16
\end{verbatim}
To generate such statement, one needs to create a name
(e.g. \texttt{\%c1}) for the location, and to identify the
corresponding LLVM type (e.g. \texttt{i16}). This is the role of the
last translation function in this section. It produces the desired
source code text and enriches the translation dictionary with the
association between the B0 variable and the created LLVM identifier. It
relies upon an auxiliary translation function $\trad{}{type}$ to be
defined hereafter:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{local} : & \textsf{ Node } \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{local} \isdef & 
  \textbf{ let } T_v = \fresh(\texttt{"\%"}\conc\textsf{v.ident}, D) \textbf{ in}  \\
  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v} \mapsto T_v \}$}{local} \textbf{ in}  \\
  & \textbf{ yields } T_v \conc \texttt{" = alloca "} \conc \trad{v.type}{type} \conc \nl \conc T_n, D_n  \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Var}  \\
  \trad{$\bot$, $D$}{local} \isdef & 
  \textbf{ yields } \mty, D 
\end{split}
\end{align}
\end{framed}

@@

@@

Suppose a machine $M$ is seen from the implementation being translated
to LLVM. For each constant, set, variable, and operation found in $M$,
the LLVM code must include a corresponding declaration. As an
illustration, consider the following C code:
\begin{verbatim}
extern int f(int, int, int *);
extern char const k1;
extern int k2;
\end{verbatim}
The corresponding LLVM produced by the C compiler is:
\begin{verbatim}
declare i32 @f(i32, i32, i32*)
@k1 = external constant i8
@k2 = external global i32
\end{verbatim}
Also, for the following C++ code:
\begin{verbatim}
class t1;
extern t1 c1;
\end{verbatim}
the compiler produces the following LLVM:
\begin{verbatim}
%class.t1 = type opaque
@c1 = external global %class.t1
\end{verbatim}

>>>>

\begin{table}
  \begin{center}
    \begin{tabular}[t]{cc}
      \begin{tabular}[t]{|l|l|l|}
        \hline
        \textsf{k} & Decription & attributes\\
        \hline
        \hline
        \textsf{Impl} & Implementation & $\textsf{sees} : \textsf{Identifiers} (?)$\\
        & & $\textsf{imports} \in \textsf{Identifiers} (?)$ \\
        & & $\textsf{values} \in \textsf{Identifiers} (?)$ \\
        & & $\textsf{vars} \in \textsf{Identifiers} (?)$ \\
        & & $\textsf{initialization} \in Inst (?)$ \\
        & & $\cdots$ \\
        & & \textsf{ops} : Op \\
        \hline
        \textsf{Op} & Operation & $\textsf{header} \in \textsf{Hd} $\\
        & & $\textsf{body} \in \textsf{inst}$ \\
        & & $\textsf{next} \in \textsf{Op} $\\
        \hline
        \textsf{Hd} & Operation header & $ \textsf{ident} \in text$ \\
        & & $\textsf{output} \in \textsf{VarDecl}$ \\
        & & $\textsf{input} \in \textsf{VarDecl}$ \\
        \hline
        \textsf{Var} & Variable, parameter & $ \textsf{ident} \in text $\\
        & & $\textsf{type} \in basic type $\\
        & & $\textsf{next} \in \textsf{Var} $\\
        \hline
      \end{tabular}
      &
      \begin{tabular}[t]{|l|l|l|}
        \hline
        \textsf{k} & Decription & attributes\\
        \hline
        \hline
        \textsf{Blk} & Block instr. & $\textsf{body} \in \textsf{inst}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{VarDecl} & Variable decl. & $\textsf{ident}\in \textsf{VarDecl} $\\
        & & $\textsf{body} \in \textsf{inst}$ \\
        \hline
        \textsf{If} & If & $\textsf{branches} \in \textsf{IfBr}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{IfBr} & If branch & $ \textsf{cond}\in predicate $ \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & $\textsf{next} \in \textsf{IfBr}$ \\
        \hline
        \textsf{BEq} & Becomes equal & $\textsf{lhs} \in target $ \\
        & & $\textsf{rhs} \in expression $ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Call} & Call instruction & $\textsf{fun} \in name$ \\
        & & $\textsf{args} \in expression$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{While} & While instr. & $\textsf{cond} \in predicate $\\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Assert} & Assert instr. & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Case} & Case instr. & $\textsf{expr} \in expression$\\
        & & $\textsf{branches} \in \textsf{CaseBr}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{CaseBr} & Case branch & $\textsf{val} \in \textsf{Value}$ \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & $\textsf{next} \in \textsf{CaseBr}$ \\
        \hline
        \textsf{Value} & Value & ??? \\
        & & \textsf{next}: \textsf{Value} \\
        \hline
      \end{tabular} \\
      \\
      \multicolumn{2}{c}{Note: $\textsf{Instr} = \{\textsf{Blk}, \textsf{VarDecl}, \textsf{If}, \textsf{BEq}, \textsf{Call}, \textsf{While}, \textsf{Assert}, \textsf{Case}\}$} \\
    \end{tabular}
    \caption{Structure of the parse tree nodes. For each attribute we give the name of the attribute and
    the type or kind(s) of node that it may be.}
    \label{tab:node-attr}
  \end{center}
\end{table}

Code generation is defined through a set of numbered rules. The
translation rules for some element \textsf{E} of the B0 language are
presented using a pseudo-functional language.  In such rules, the
\textsf{sans serif} and \texttt{typewriter} fonts are used to denote
\textsf{B0} abstract syntax tree nodes and \texttt{LLVM} text
respectively, and $\conc$ denotes string catenation and $\nl$ a new
line.

@@

$\fresh: text \times dict \rightarrow text$ is such that
$\fresh(t, D)$  returns a text $t'$ that is either $t$, if $t$ is not in
the domain of the dictionary $D$, or $t$ followed by the string $n$ of
digits representing the smallest positive number $t \conc n$ does not
belong to the domain of $D$. This dictionary is a function that keeps 
\textbf{B identifiers} in domain and \textbf{LLVM identifiers} in range.

@@

We first define the translation rule for the root of the source
program abstract syntax tree, which shall correspond to a B
implementation. This is a preliminary version where the code
generation is only able to handle standalone B0 implementations, i.e.
the state is defined in the ``concrete variables'' clause. Moreover
the data is restricted to predefined basic types \textsf{INT} and
\textsf{BOOL}:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
  \trad{i}{impl} \isdef & \textbf{ let } Dic = \{ \#ModuleName \mapsto  i.moduleName\} \textbf{ in}\\
  & \textbf{ let } O_{c}, Dic'  = \trad{i.values, Dic }{val} \textbf{ and } O_v, Dic'' = \trad{i.vars, Dic'}{var} \\
  & \quad \textbf{ and } O_{in} = \trad{i.initialization, Dic''}{ini} \textbf{ and } O_{op} = \trad{i.ops, Dic''}{ops} \\
  & \quad \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad O_{c} \conc O_{v} \conc O_{in} \conc O_{op}    \\
  & \textbf{ where } \textsf{i.k} = \textsf{Impl} 
\end{split}
\end{align}
\end{framed}

Code generation for initialisation and operations is specified in
section~\ref{sec:control}. Code generation for data (type and
state) is defined in section~\ref{sec:data}, and for expression
evaluation in section~\ref{sec:expr}.

@@

\begin{framed}
\begin{align}
\begin{split}
  \trad{}{var} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{var} \isdef & textListVar = \trad{v,$D$,0}{DicVars}  \textbf{ in}\\
  & \textbf{ yields } \\
  & \quad \texttt{" \%struct."}\conc  D(\#ModuleName) \conc \texttt{"\_internalState = type \{"} \\
  & \quad  \conc textListVar \conc \texttt{"\} "}  \nl , D \\
  & \textbf{ where } \textsf{v} \ne  \bot \textbf{ and }  \textsf{v.k} = ConcreteVars \\
  \trad{$\bot$, $D$}{var} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

The type list of module variables are declared using this role, where:
n is the actual node, D is the dictionary and  p is the position of variable.

\begin{framed}
\begin{align}
\begin{split}
  \trad{}{DicVars} : & \textsf{ Node } \times dict \times integer \rightarrow text  \times dict \\
  \trad{n, $D$, p }{DicVars} \isdef  & \textbf{ let }   type = \trad{n.type}{type}   \textbf{ and }  t', D' =  \trad{n.next,D,p+1} {DicVars}   \textbf{ in}   \\
  & \textbf{ yields } \\
  & \quad  type \conc^,  t' , D' \cup \{ \textsf{n.ident} \mapsto (\textit{text }  p) \}  \cup\\
  & \quad \{ \texttt{"\#type\_"}\conc \textsf{n.ident} \mapsto type \} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Var}  \\
\trad{$\bot$, $D$, p }{DicVars} \isdef & \textbf{yields} \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

Constant valuations are translated to constants in the
LLVM program, as described in the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{val} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{val} \isdef & \textbf{ let } type = \trad{n.type}{type}   \textbf{ and }O', D' = \trad{v.next, $D$}{val} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = constant "} \conc type \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{v.ident} \mapsto id \} \conc  \{ \texttt{"\#type\_"}\conc \textsf{n.ident} \mapsto type \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Values} \\
  \trad{$\bot$, $D$}{val} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

@@

This section addresses how B operations are translated to LLVM
functions. Each operation is translated in turn, and the result of
the translation of individual operations is concatenated to yield the
result of the translation of the operations clause:
\begin{framed}
\begin{align}
  \begin{split}
  \trad{}{ops} : & \textsf{ Node} \times \textsf{ Dict} \rightarrow text \\
  \trad{o,D}{ops} \isdef 
  & \textbf{ yields } \trad{o,D}{op} \conc \trad{o.next,D}{ops}  \\
  & \textbf{ where } \textsf{o} \neq \bot  \textbf{ and } \textsf{o.k} \in \textsf{Op}  \\
  \trad{$\bot$,D}{ops} \isdef
  & \textbf{ yields } \mty 
  \end{split}
\end{align}
\end{framed}

@@

The translation of an operation is defined by the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{op} &: \textsf{ Node } \times \textsf{ Dict} \rightarrow text \\
  \trad{op,d}{op} \isdef
  & \textbf{ let } O_l = \trad{op}{link} \textbf{ and } O_h, D_h = \trad{op.header,d}{header} \textbf{ in}  \\
  & \textbf{ let } O_a, D_a = \trad{op.body, $D_h$}{alloc} \textbf{ in}  \\
  & \textbf{ let } O_i = \trad{op.body, $D_a$}{inst} \textbf{ in}  \\
  & \textbf{ yields } \texttt{"define "} \conc O_l \conc O_h \conc \texttt{"\{"} \conc \nl \conc \\
  & \quad \conc O_a \conc \nl \conc O_i \conc \nl \conc \texttt{"ret void"} \conc \nl \conc \texttt{"\}"} \conc \nl  \\
  & \textbf{ where } \textsf{op} \ne \bot \textbf{ and } \textsf{op.k} = \textsf{Op} 
\end{split}
\end{align}
\end{framed}

@@

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{header} &: Hd \times dict \rightarrow text \times dict \\
    \trad{h,D}{header} \isdef& \textbf{ let } T_n = \textsf{"@"} \conc \textsf{h.ident}  \textbf{ and } 
    \textbf{ and } T_i, D_i = \trad{h.input}{inp} \\
    & \textbf{ and } T_o, D_o = \trad{h.output}{out} \textbf{ in}  \\
    & \textbf{ let } T_p = \texttt{" \%struct."} \conc D(\#ModuleName) \conc \\
    & \texttt{"\_internalState* nocapture \%pointerMVars"}  \textbf{ in}  \\
    & \textbf{ yields } \texttt{"void "} \conc T_n \conc \texttt{"("} \conc T_p \conc^, T_i \conc^, T_o \conc
    \texttt{")"}, D_i \cup D_o \conc \{ \textsf{h.ident} \mapsto T_n \}  \\
    & \textbf{ where } \textsf{h} \ne \bot \textbf{ and } \textsf{h.k} = \textsf{Hd} 
  \end{split}
\end{align}
\end{framed}

@@

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{inp} &: \textsf{ Node } \rightarrow text \times dict \\
    \trad{i}{inp} \isdef& \textbf{ let } T_n = \textsf{"\%"} \conc \textsf{i.ident} 
    \textbf{ and } T_t = \trad{i.type}{type} 
    \textbf{ and } T, D = \trad{i.next}{inp} \textbf{ in}  \\
    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{i.ident} \mapsto T_n \} \cup D  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Var}  \\
    \trad{$\bot$}{inp} \isdef& \textbf{ yields } \mty, \emptyset 
  \end{split}
\end{align}
\end{framed}

@@

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{out} &: \textsf{ Node } \rightarrow text \times dict \\
    \trad{o}{out} \isdef& \textbf{ let } T_n = \textsf{"\%"} \conc \textsf{o.ident} 
    \textbf{ and } T_t = \trad{o.type}{type} \conc \texttt{"*"}
    \textbf{ and } T, D = \trad{o.next}{out} \textbf{ in}  \\
    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{o.ident} \mapsto T_n \} \cup D  \\
    & \textbf{ where } \textsf{o} \ne \bot \textbf{ and } \textsf{o.k} = \textsf{Var}  \\
    \trad{$\bot$}{out} \isdef& \textbf{ yields } \mty, \emptyset  
  \end{split}
\end{align}
\end{framed}

@@

One must avoid generating name conflicts in the LLVM output. For each
B0 variable \textsf{v} found, a fresh LLVM name \texttt{l} is thus
created. This association is recorded in the $D$ parameters and results.
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{alloc}_{base} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ yields } \trad{i.body, $D$}{alloc}  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{Blk}, \textsf{While}, \textsf{IfBr}, \textsf{CaseBr} \}  \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ yields } \trad{i.branch, $D$}{alloc}  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{If}, \textsf{Case} \}  \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ let } O_l, D_l = \trad{i.ident, $D$}{local} \textbf{ in}  \\
    & \textbf{ let } O_b, D_b = \trad{i.body, $D_l$}{alloc} \textbf{ in}  \\
    & \textbf{ yields } O_l \conc \nl \conc O_b, D_b  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{VarDecl} \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ let } O_l, D_l = \trad{i.ident, $D$}{global} \textbf{ in}  \\
    & \textbf{ let } O_b, D_b = \trad{i.body, $D_l$}{alloc} \textbf{ in}  \\
    & \textbf{ yields } O_l \conc \nl \conc O_b, D_b  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Var} 
  \end{split}
\end{align}
\end{framed}

This function get internal references to concrete variables from module.

\begin{framed}
\begin{align}
\begin{split}
  \trad{}{global} : & \textsf{ Node } \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{global} \isdef & 
  \textbf{ let } T_v = \fresh(\texttt{"\%"}\conc\textsf{v.ident}, D) \textbf{ in}  \\
  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v.ident} \mapsto T_v \}$}{local} \textbf{ in}  \\
  & \textbf{ yields } T_v \conc \texttt{" = getelementptr inbounds "} \conc\\
  & \texttt{"\%struct.ModuloName\_internalState* \%pointerMVars, i64 0, i32 "} \conc \\
  & D(\texttt{"\#type\_"}\conc v.ident) \conc \nl \conc T_n, D_n  \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Var}  \\
  \trad{$\bot$, $D$}{global} \isdef & 
  \textbf{ yields } \mty, D 
\end{split}
\end{align}
\end{framed}

To illustrate how heap allocation works in LLVM, this is the statement
produced by Clang corresponding to the C declaration
\verb^unsigned short c1^:
\begin{verbatim}
  %c1 = alloca i16
\end{verbatim}
To generate such statement, one needs to create a name
(e.g. \texttt{\%c1}) for the location, and to identify the
corresponding LLVM type (e.g. \texttt{i16}). This is the role of the
last translation function in this section. It produces the desired
source code text and enriches the translation dictionary with the
association between the B0 variable and the created LLVM identifier. It
relies upon an auxiliary translation function $\trad{}{type}$ to be
defined hereafter:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{local} : & \textsf{ Node } \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{local} \isdef & 
  \textbf{ let } T_v = \fresh(\texttt{"\%"}\conc\textsf{v.ident}, D) \textbf{ in}  \\
  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v.ident} \mapsto T_v \}$}{local} \textbf{ in}  \\
  & \textbf{ yields } T_v \conc \texttt{" = alloca "} \conc \trad{v.type}{type} \conc \nl \conc T_n, D_n  \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Var}  \\
  \trad{$\bot$, $D$}{local} \isdef & 
  \textbf{ yields } \mty, D 
\end{split}
\end{align}
\end{framed}

@@

B method supported different types of modularity.
The implementation modules can use the clauses: SEES, PROMOTES and IMPORT.

@@

Suppose a machine $M$ is seen from the implementation being translated
to LLVM. For each constant, set, variable, and operation found in $M$,
the LLVM code must include a corresponding declaration. As an
illustration, consider the following C code:
\begin{verbatim}
extern int f(int, int, int *);
extern char const k1;
extern int k2;
\end{verbatim}
The corresponding LLVM produced by the C compiler is:
\begin{verbatim}
declare i32 @f(i32, i32, i32*)
@k1 = external constant i8
@k2 = external global i32
\end{verbatim}

[[TODO: to add and to explain the translation function]]

\subsection{The IMPORT clause}
